Relocating a Cluster of Earthquakes Using a Single Seismic Station

### 2012-12-12 paper2/paper2.tex
incorporated most of Juerg's comments + started a second pass of the other review comments. I need to complete the second pass of Trevor Allen's comments then update figures and decided whether to add the suggested references

\title{Relocating a cluster of earthquakes using a single station}
\author[D.J. Robinson, M. Sambridge, R. Snieder and J Hauser]
  {David J. Robinson$^{1,2}$, Malcolm Sambridge$^{1}$, Roel Snieder$^{3}$, J\"urg Hauser$^{4}$ \\
  $^1$ Research School of Earth Sciences, Australian National
    University, Canberra ACT \emph{0200}, Australia, E-mail: david.robinson@anu.edu.au \\
$^2$  Risk Analysis Methods, Geoscience Australia, GPO Box \emph{383} Canberra ACT \emph{2601} Australia, E-mail: david.robinson@ga.gov.au\\
$^3$ Center for Wave Phenomena and Department of Geophysics, Colorado School of Mines, Golden CO \emph{80401-1887}, USA \\
$^4$ ..................}

@@

Earthquake location is important for many applications. Locations
are required for:  magnitude
determination \citep{dr_Richter35a, dr_Gutenberg45a};
computing moment tensors \citep{dr_Sipkin02a};
seismological studies of the Earth's interior
\citep{dr_Spencer80a, dr_Kennett95a, dr_Curtis02a, dr_Kennett04a};
understanding strong motion and seismic attenuation
\citep{dr_Toro97a, dr_Campbell03a}
 and modeling earthquake hazard or risk
\citep{dr_Frankel00a, dr_Stirling02a, dr_Robinson06b}.
The accuracy required in earthquake location depends on the
application. For example, imaging the structure of a fracture system
from microseismicity requires greater detail than determining
whether a $M_w=7.5$ earthquake occurs offshore for tsunami warning.
 This paper focuses on reducing location uncertainty for
a cluster of events when they are recorded by a small
number of stations.

Absolute location describes the location of an earthquake with
respect to a global reference such as latitude, longitude (or
easting/northing) and depth. Uncertainties associated with absolute
locations are influenced by source to station distances, the number
of stations and their geometry, quality of signal and accuracy of
the velocity model used in computing travel times. Uncertainties in
absolute location are typically of the order of several kilometers
because they are susceptible to uncertainty in the velocity
structure along the entire path between the source and receiver. For
example, \citet{dr_Shearer99a} states that location uncertainty in
the ISC and PDE catalogues are generally around 25\,km horizontally
and at least 25\,km in depth.\footnote{Here the depth uncertainties
of 25\,km assume the use of depth dependent phases such as $pP$.
Without such phases the uncertainty is higher.} \citet{dr_Bondar04a}
demonstrate that at the local scale, absolute locations are accurate
to within 5\,km with a 95\% confidence level when local networks
meet a number of station related criteria. Such errors are too large
for many applications, particularly those focussed on imaging
rupture surfaces from aftershock sequences.

@@

In traditional approaches to absolute and relative location only
early onset body waves, typically $P$ and/or $S$ waves, are used.
The data utilised may be the direct arrival times; travel time
difference computed between picked arrivals of two waveforms; or
time differences inferred from time-lagged cross correlation of
relatively small windows around the body wave arrivals.
 In all three cases, the majority of the
waveform is discarded. Furthermore, obtaining high accuracy with these techniques
requires multiple stations with good azimuthal coverage.
In this paper we demonstrate that it is possible to
significantly reduce location uncertainty when few stations are
available by using more of the waveform.

Coda refers to  later arriving waves in the seismogram that arise
from scattering  \citep{dr_Aki69a,dr_Snieder99a,dr_Snieder06a}. Coda
waves are ignored in most seismological applications due to the
complexity involved in modeling their generation. In this paper we
develop an approach for locating earthquakes using coda waves.
\citet{dr_Snieder05a} demonstrate that the coda of two earthquakes
can be used to estimate the separation between them. Their
technique, known as coda wave interferometry (CWI), is based on the
interference pattern between the coda waves. Unlike travel time
techniques, CWI
 does not require multiple stations or good azimuthal coverage.
In fact, it is possible to obtain estimates of separation using a single
station \citep{dr_Robinson07b}. This makes CWI particularly interesting
for regions where station density is low such as intraplate settings. In this
paper we demonstrate how CWI separation estimates can be used to constrain
location with data from a single station.
Our technique can be used on coda waves alone or in combination with
travel times.

@@

\citet{dr_Robinson11a} show that coda waves provide only
probabilistic constraints on source separation and introduce a
Bayesian approach for describing the probability of true separation
given the CWI data. Their approach is summarised by
\begin{equation}
\label{eq:-bayesian-noisyform}
P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN}) \propto P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)
\times P(\widetilde{\delta}_t)
\end{equation}
where  $P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$ is the posterior function
indicating the probability of true separation $\widetilde{\delta}_t$ given the noisy CWI separation
estimates $\widetilde{\delta}_{CWIN}$; $P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)$
is the likelihood function (or forward model) giving the probability that the separation
estimates $\widetilde{\delta}_{CWIN}$ would be observed if the true separation was $\widetilde{\delta}_t$; and
$P(\widetilde{\delta}_t)$ is the prior PDF accounting for all a-priori information.
The tilde above the separation parameters in eq. (\ref{eq:-bayesian-noisyform})
indicates the use of a wavelength normalised separation parameter
\begin{equation}
\label{eq-normalisation-eqn}
\widetilde{\delta} = \frac{\delta}{\lambda_{d}},
\end{equation}
which measures separation ($\delta$ = $\delta_{CWIN}$ or $\delta_t$)
with respect to dominant wavelength $\lambda_{d}$. In this paper we
consider a uniform prior over appropriate bounds to ensure that the 
posterior function is dominated by the recorded data.
The procedure for computing the noisy likelihood
$P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)$ is derived by
\citet{dr_Robinson11a} and summarised in Appendix
\ref{sec-Appendix-noisylikelihood}. With these two pieces in place
we can compute the posterior
$P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$ (or PDF) for the
separation between any pair of events directly from their coda
waves.

We seek a probability density function (PDF) which links
 individual pairwise posteriors $P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$
to describe the location of multiple events whose maximum corresponds to
 the most probable combination of locations.
More importantly however, the PDF shall
quantify location uncertainty and provide information
on the degree to which individual events are constrained
 by the data. For convenience, we begin with three earthquakes having locations
$\mathbf{e}_1$, $\mathbf{e}_2$ and $\mathbf{e}_3$. Using a Bayesian
formulation we write
\begin{equation}
\label{eq-Bayes-three-locations}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3| \mathbf{d}) \propto P( \mathbf{d}|\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)
\times P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3),
\end{equation}
where $P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \mathbf{d})$,
$P(\mathbf{d}|\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ and
$P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ are the posterior,
likelihood and prior functions, respectively. In equation
(\ref{eq-Bayes-three-locations}) $\mathbf{d}$ represents
observations that constrain the locations. They can be any
combination of travel times, geodetic information  or CWI
separations. For example, if coda waves are used we have
$P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 |
\widetilde{\mitbf{\delta}}_{CWIN})$ and
$P(\widetilde{\mitbf{\delta}}_{CWIN}|\mathbf{e}_1, \mathbf{e}_2,
\mathbf{e}_3)$ where $\widetilde{\mitbf{\delta}}_{CWIN}$ are the
wavelength normalised separation estimates. Alternatively, if we use
CWI and travel time data we may write $P(\mathbf{e}_1, \mathbf{e}_2,
\mathbf{e}_3 | \widetilde{\mitbf{\delta}}_{CWIN},
\mathbf{\Delta}_{TT})$ and
$P(\widetilde{\mitbf{\delta}}_{CWIN},\mathbf{\Delta}_{TT}|\mathbf{e}_1,
\mathbf{e}_2, \mathbf{e}_3)$ where $\mathbf{\Delta}_{TT}$ represent
travel time differences. In this derivation and in Section
\ref{sec:benchmarking} we focus on the constraints imposed by coda
waves, whereas in Section \ref{sec:CalaverasLoc-CWIandTT} we
demonstrate how CWI and travel time data can be combined.

@@

The event locations $\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n$
are defined by coordinates $\hat{x}$, $\hat{y}$ and $\hat{z}$
where the hat $\hat{.}$ indicates use of
a local coordinate system. We choose a local coordinate system
which removes ambiguity associated with transformations of the coordinate system. It is 
necessary to do this because the distances between events are invariant for rotations
and translations of the seismicity patterns and hence we know in advance that rotations 
and translations can not be resolved from CWI alone.  
In defining this coordinate system we fix the first event at the origin,
\begin{equation}
\label{eq:coord-const1}
\mathbf{e}_1 = (0,0,0),
\end{equation}
the second event on the positive $\hat{x}$-axis
\begin{equation}
\label{eq:coord-const2}
\mathbf{e}_2 = (\hat{x}_2, 0, 0), \hat{x}_2>0
\end{equation}
the third on the $\hat{x}-\hat{y}$ plane
\begin{equation}
\label{eq:coord-const3}
\mathbf{e}_3 = (\hat{x}_3,\hat{y}_3,0), \hat{y}_3>0
\end{equation}
and the fourth to
\begin{equation}
\label{eq:coord-const4}
\mathbf{e}_4 = (\hat{x}_4,\hat{y}_4,\hat{z}_4), \hat{z}_4>0.
\end{equation}
This coordinate system reduces translational (constraint \ref{eq:coord-const1})
and rotational (constraints \ref{eq:coord-const2} to \ref{eq:coord-const4}) non-uniqueness
without loss of generality. It is necessary to work with a local coordinate
system when using coda waves alone because the CWI technique constrains only event separation
between earthquakes. The inclusion of travel times in Section \ref{sec:CalaverasLoc-CWIandTT}
allows us to move to a global reference system.

In summary, the posterior $P(\mathbf{e}_1,...,\mathbf{e}_n |
\widetilde{\mathbf{\delta}}_{CWIN})$ and its negative logarithm $L$
describe the joint probability of multiple event locations given the
observed coda waves. The most likely set of locations is given by
the minimum of $L$.   In this paper we use the Polak-Ribiere
technique \citep{dr_Press87a}, a conjugate gradient method, to
optimize $L$. It uses the derivatives of $L$, derived in Appendix
\ref{sec-Appendix-derivatives_ofL}, to guide the
optimization procedure. Note that when optimizing
equation \ref{eq:-negative-log} the values of $ln \left[ c \right]$
and $ln \left[ P(e_i) \right]$ can be ignored because they are
constant\footnote{$ln \left[ P(e_i) \right]$ is constant because we
consider a uniform prior.}

@@

The local coordinate system works by constraining the location of
the first three earthquakes. Earthquake 1 is fixed at the origin,
earthquake 2 on the positive $\hat{x}$-axis and earthquake 3 has
$\hat{y}>0$. As the number of events increase the strength of these
constraints on later events weakens allowing small rotations of
events with respect to each other. That is, even though the
rotational freedom of the cluster is in principal removed by the
constraints imposed on the events (see equations
(\ref{eq:coord-const1}) to (\ref{eq:coord-const3})\footnote{equation
\ref{eq:coord-const4} is needed in 3D only}) we observe that in
practice the presence of noise allows the rotational non-uniqueness
to reappear. This is because the `easiest' way data noise can
propagate is into the direction which is least constrained by the
data. The same phenomena is observed in linear inversion where noise
creates large spurious model changes in directions of the
eigenvectors with the smallest singular values \citep{dr_Aster05a}.
 Fortunately however, combining coda waves with
measurements of travel times alleviates this problem and facilitate
the removal of a local coordinate system altogether (see Section
\ref{sec:CalaverasLoc-CWIandTT}).

@@

Figures \ref{fig:optimisationresults-2Dsynth}(a) and (b) illustrates
the maximum $\Delta_{max}$ (top) and mean $\Delta_\mu$ (middle) of
the coordinate error as a function of increasing linkage. We show
the statistics for the `best' optimization solution (black) and for
the solution space when all 25 optimizations are considered (gray).
In the latter case the best solution is determined by the set of
event locations which lead to the smallest value of $L$. The error
in the best solution is consistent when 30\% or more of the branches
are used. The errors increase when only 10\% or 20\% of the
constraints are included. Interestingly, this breakdown around 20\%
to 30\% coincides with the point where the average number of
branches required to link each event pair reaches 2 (see
Fig.\ref{fig:optimisationresults-2Dsynth} (bottom)). Since the
average number of branches can be computed in advance it can be used
as an indication of the inversion stability prior to optimization. A
higher breakdown is observed when all 25 solutions are considered
collectively. For example, the maximum coordinate error
$\Delta_{max}$ exceeds that for the best solution for linkage
$\leq$60\% confirming that the optimization is susceptible to local
minima and that a range of starting points should be considered.
Some optimizations fail to converge after 1200 iterations when the
linkage is 60\% or lower. All optimizations fail when then linkage
is 20\% or lower. Despite their failure to converge, the locations
at final iteration are close to the actual solution.

The derivatives used in the conjugate gradient method depend on
events connected by CWI measurements. Consequently, earthquakes that
are only connected via other events do not `communicate' with each
other directly. To some extent, this should be adressed during the
iterative process where location information can spread to events
which have no direct links. However, the lack of direct connection
through the gradient could prevent convergence in extreme cases, or
more likely slow the procedure down. This could explain why some
examples do not converge after 1200 iterations.
\citet{dr_VanDecar94a} show that gradient damping acts slowly
through iterative least-squares, because
 every cell in one iteration communicates only with its neighbours, and they demonstrate that this can be
fixed with preconditioning in some cases. Their findings suggest
that it may be possible to improve the convergence (stability and/or
speed) of the CWI optimization by preconditioning.

@@

When 70\% of the direct constraints are considered all optimization
results (gray) are consistent with the best solution (black). The
best solution constrains the event locations down to 30\% of the
direct links. There is one notable difference between the 3D and 2D results. In 2D
the final iteration was close to the actual solution when the
optimization failed to converge. Conversely, in 3D the optimization
appears to converge to the correct solution or fail completely,
leading to a set of locations at final iteration which do not
resemble the actual solution. This is depicted in Figure
\ref{fig:optimisationresults-3Dsynth} by the absence of the gray and
black lines below 60\% and 30\% of the constraints, respectively.
The reason for this difference may be due to the increased number of
degrees of freedom in 3D requiring a greater number of iterations to
converge. Nevertheless, the accurate convergence of the best
solution for cases with 30\% linkage or higher is encouraging for
the potential of coda wave optimization to constrain earthquake
location.

@@

In summary, the synthetic examples demonstrate the ability of coda
wave data to constrain relative event location using optimization.
The optimization error is influenced by the noise on CWI estimates
with greater $\bar{\sigma}_N$ leading to larger errors in the
solutions. When 70\% or more of the direct branches are used the
optimiser is stable with no observable difference in the solution
for 25 randomly chosen starting locations. As the direct linkage
reduces to 50\% the optimization becomes less stable and the best
solution from 25 random starting locations is required to find the
optimal solution. As the number of links decrease below 30\% the
best solution fails to converge.

@@

\begin{figure}
\noindent\includegraphics[width =
20pc]{diags/CalaverasMap/gmt_california/CaliforniaCalaverasMap1.eps}
\caption{Elevation in California showing location of the Calaveras
cluster (white star) and 805 seismic stations (black triangles).}
\label{fig:-eqopti-California-Calaveras}
\end{figure}

@@

Global locations cannot be found by CWI alone. For the sake of
comparison, we arbitrarily choose a `master' event and translate our
relative locations to align with the hypoDD location for the same
event. This arbitrary translation does not change the relative
locations. We return to this issue of relative versus actual
location in Example 7 by introducing a combined travel time and coda
wave inversion.

@@

% This is the version using hypoDD with SVD
\begin{figure}
\noindent\includegraphics[width =
20pc]{diags/CalaverasLoc4_hypoDD_SVD.eps} \caption{Example 6 -
Statistics on coordinate differences for reduced station inversions.
Differences are computed between the inversion results (CWI and
hypoDD) and the complete hypoDD locations for all 308 events. The
top subplot illustrates the number of constrainable events,$nE$ in
the CWI and hypoDD inversions as a function of the stations
considered.} \label{fig-statremoval_summarystats}
\end{figure}
%%=========================================================================

The hypoDD algorithm also struggles to locate all events as the number
of stations is reduced. In the case of hypoDD an event can be
identified as unconstrainable in one of two stages. Firstly, the
data are analysed to ensure that there exists travel time
differences for each event and at least one other earthquake. This
is analogous to the situation for the coda wave technique. The
hypoDD program also has a secondary identification phase in which
events that can not be located sufficiently are rejected during the
inversion. This process is presumably related to the iterative
removal of outliers described by \citet{dr_Waldhauser00a}. The top
panel of Figure \ref{fig-statremoval_summarystats} shows that the
number of events re-located by hypoDD fluctuates between 63 and 28
earthquakes for 10 to 4 stations and it demonstrates that the number
of events located by hypoDD is less than the number located by CWI
in all cases.

>>>>

\title{Relocating a Cluster of Earthquakes Using a Single Seismic Station}
\author[D.J. Robinson, M. Sambridge, R. Snieder and J Hauser]
  {David J. Robinson$^{1,2}$, Malcolm Sambridge$^{1}$, Roel Snieder$^{3}$, Juerg Hauser$^{4}$ \\
  $^1$ Research School of Earth Sciences, Australian National
    University, Canberra ACT \emph{0200}, Australia, E-mail: david.robinson@anu.edu.au \\
$^2$  Risk Analysis Methods, Geoscience Australia, GPO Box \emph{383} Canberra ACT \emph{2601} Australia, E-mail: david.robinson@ga.gov.au\\
$^3$ Center for Wave Phenomena and Department of Geophysics, Colorado School of Mines, Golden CO \emph{80401-1887}, USA \\
$^4$ CSIRO Earth Science and Resource Engineering, Dick Perry Av, Kensington WA \emph{6151} Australia}

@@

Accurate earthquake location is important for many applications. Locations
are required for:  magnitude
determination \citep{dr_Richter35a, dr_Gutenberg45a};
computing moment tensors \citep{dr_Sipkin02a};
seismological studies of the Earth's interior
\citep{dr_Spencer80a, dr_Kennett95a, dr_Curtis02a, dr_Kennett04a};
understanding strong motion and seismic attenuation
\citep{dr_Toro97a, dr_Campbell03a}
 and modeling earthquake hazard or risk
\citep{dr_Frankel00a, dr_Stirling02a, dr_Robinson06b}.
The accuracy required in earthquake location depends on the
application. For example, imaging the structure of a fracture system
from microseismicity requires greater detail than determining
whether a $M_w=7.5$ earthquake occurs offshore for tsunami warning.
 This paper focuses on reducing location uncertainty for
a cluster of events when they are recorded by a small
number of stations.

Absolute location describes the location of an earthquake with
respect to a global reference such as latitude, longitude (or
easting/northing) and depth. Uncertainties associated with absolute
locations are influenced by source to station distances, the number
of stations and their geometry, signal-to-noise, clarity of onsets and accuracy of
the velocity model used in computing travel times. Uncertainties in
absolute location are typically of the order of several kilometers
because they are susceptible to uncertainty in the velocity
structure along the entire path between the source and receiver. For
example, \citet{dr_Shearer99a} states that location uncertainty in
the ISC and PDE catalogues are generally around 25\,km horizontally
and at least 25\,km in depth.\footnote{Here the depth uncertainties
of 25\,km assume the use of depth dependent phases such as $pP$.
Without such phases the uncertainty is higher.} \citet{dr_Bondar04a}
demonstrate that at the local scale, absolute locations are accurate
to within 5\,km with a 95\% confidence level when local networks
meet a number of station related criteria. Such errors are too large
for many applications, particularly those focussed on imaging
rupture surfaces from aftershock sequences.

@@

In traditional approaches to absolute and relative location only
early onset body waves or crustal phases are used.
The data utilised may be the direct arrival times; travel time
difference computed between picked arrivals of two waveforms; or
time differences inferred from time-lagged cross correlation of
relatively small windows around the body wave arrivals.
 In all three cases, the majority of the
waveform is discarded. Furthermore, obtaining high accuracy with these techniques
requires multiple stations with good azimuthal coverage.
In this paper we demonstrate that it is possible to
significantly reduce location uncertainty when few stations are
available by using more of the waveform.

Coda refers to  later arriving waves in the seismogram that arise
from scattering  \citep{dr_Aki69a,dr_Snieder99a,dr_Snieder06a}. Coda
waves are ignored in most seismological applications due to the
complexity involved in constraining complex hetergeneous velocity models 
in real settings. In this paper we
develop an approach for locating earthquakes using coda waves.
\citet{dr_Snieder05a} demonstrate that the coda of two earthquakes
can be used to estimate the separation between them. Their
technique, known as coda wave interferometry (CWI), is based on the
interference pattern between the coda waves. Unlike travel time
based location techniques, CWI
 does not require multiple stations or good azimuthal coverage.
In fact, it is possible to obtain estimates of separation using a single
station \citep{dr_Robinson07b}. This makes CWI particularly interesting
for regions where station density is low such as intraplate settings. In this
paper we demonstrate how CWI separation estimates can be used to constrain
location with data from a single station.
Our technique can be used on coda waves alone or in combination with
travel times. We begin by introducing the theory of CWI based earthquake location 
(Section \ref{sec:theory}). This is followed by a demonstration of capability using 
synthetic examples (Section \ref{sec:benchmarking}) and application to earthquakes 
on the Calaveras fault, California using CWI alone (Section \ref{sec:CalaverasLoc-CWIonly}
and CWI in combination with travel time constraints (Section \ref{sec:CalaverasLoc-CWIandTT}).

@@

\citet{dr_Robinson11a} show that CWI leads to
probabilistic constraints on source separation and introduce a
Bayesian approach for describing the probability of true separation
given the CWI data. Their approach is summarised by
\begin{equation}
\label{eq:-bayesian-noisyform}
P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN}) \propto P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)
\times P(\widetilde{\delta}_t)
\end{equation}
where  $P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$ is the posterior function
indicating the probability of true separation $\widetilde{\delta}_t$ given the noisy CWI separation
estimates $\widetilde{\delta}_{CWIN}$; $P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)$
is the likelihood function (or forward model) giving the probability that the separation
estimates $\widetilde{\delta}_{CWIN}$ would be observed if the true separation was $\widetilde{\delta}_t$; and
$P(\widetilde{\delta}_t)$ is the prior PDF accounting for all a-priori information.
The tilde above the separation parameters in eq. (\ref{eq:-bayesian-noisyform})
indicates the use of a wavelength normalised separation parameter
\begin{equation}
\label{eq-normalisation-eqn}
\widetilde{\delta} = \frac{\delta}{\lambda_{d}},
\end{equation}
which measures separation ($\delta$ = $\delta_{CWIN}$ or $\delta_t$)
with respect to dominant wavelength $\lambda_{d}$. In this paper we
consider a uniform prior over appropriate bounds to ensure that the 
posterior function is dominated by the recorded data.
The procedure for computing the likelihood
$P(\widetilde{\delta}_{CWIN}|\widetilde{\delta}_t)$ is derived by
\citet{dr_Robinson11a} and summarised in Appendix
\ref{sec-Appendix-noisylikelihood}. With these two pieces in place
we can compute the posterior
$P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$ (or PDF) for the
separation between any pair of events directly from their coda
waves.

We seek a probability density function (PDF) which links
 individual pairwise posteriors $P(\widetilde{\delta}_t|\widetilde{\delta}_{CWIN})$
to describe the location of multiple events whose maximum corresponds to
 the most probable combination of locations.
More importantly however, the PDF shall
quantify location uncertainty and provide information
on the degree to which individual events are constrained
 by the data. For convenience, we begin with three earthquakes having locations
$\mathbf{e}_1$, $\mathbf{e}_2$ and $\mathbf{e}_3$. Using a Bayesian
formulation we write
\begin{equation}
\label{eq-Bayes-three-locations}
P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3| \mathbf{d}) \propto P( \mathbf{d}|\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)
\times P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3),
\end{equation}
where $P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 | \mathbf{d})$,
$P(\mathbf{d}|\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ and
$P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3)$ are the posterior,
likelihood and prior functions, respectively. In equation
(\ref{eq-Bayes-three-locations}) $\mathbf{d}$ represents
observations that constrain the locations. They can be any
combination of travel times, geodetic information  or CWI
separations. For example, if coda waves are used we have
$P(\mathbf{e}_1, \mathbf{e}_2, \mathbf{e}_3 |
\widetilde{\mitbf{\delta}}_{CWIN})$ and
$P(\widetilde{\mitbf{\delta}}_{CWIN}|\mathbf{e}_1, \mathbf{e}_2,
\mathbf{e}_3)$, where $\widetilde{\mitbf{\delta}}_{CWIN}$ are the
wavelength normalised separation estimates. Alternatively, if we use
CWI and travel time data we may write $P(\mathbf{e}_1, \mathbf{e}_2,
\mathbf{e}_3 | \widetilde{\mitbf{\delta}}_{CWIN},
\mathbf{\Delta}_{TT})$ and
$P(\widetilde{\mitbf{\delta}}_{CWIN},\mathbf{\Delta}_{TT}|\mathbf{e}_1,
\mathbf{e}_2, \mathbf{e}_3)$ where $\mathbf{\Delta}_{TT}$ represent
travel time differences. In this derivation and in Section
\ref{sec:benchmarking} we focus on the constraints imposed by coda
waves, whereas in Section \ref{sec:CalaverasLoc-CWIandTT} we
demonstrate how CWI and travel time data can be combined.

@@

The event locations $\mathbf{e}_1, \mathbf{e}_2, ..., \mathbf{e}_n$
are defined by coordinates $\hat{x}$, $\hat{y}$ and $\hat{z}$
where the hat indicates use of
a local coordinate system. We choose a local coordinate system
which removes ambiguity associated with transformations of the coordinate system. It is 
necessary to do this because the distances between events are invariant for rotations
and translations of the seismicity patterns and hence we know in advance that rotations 
and translations can not be resolved from CWI alone.  
In defining this coordinate system we fix the first event at the origin
\begin{equation}
\label{eq:coord-const1}
\mathbf{e}_1 = (0,0,0),
\end{equation}
the second event on the positive $\hat{x}$-axis
\begin{equation}
\label{eq:coord-const2}
\mathbf{e}_2 = (\hat{x}_2, 0, 0), \hat{x}_2>0
\end{equation}
the third on the $\hat{x}-\hat{y}$ plane
\begin{equation}
\label{eq:coord-const3}
\mathbf{e}_3 = (\hat{x}_3,\hat{y}_3,0), \hat{y}_3>0
\end{equation}
and the fourth to
\begin{equation}
\label{eq:coord-const4}
\mathbf{e}_4 = (\hat{x}_4,\hat{y}_4,\hat{z}_4), \hat{z}_4>0.
\end{equation}
This coordinate system reduces translational (equation \ref{eq:coord-const1})
and rotational (equations \ref{eq:coord-const2} to \ref{eq:coord-const4}) non-uniqueness
without loss of generality. It is necessary to work with a local coordinate
system when using coda waves alone because the CWI technique constrains only event separation
between earthquakes. The inclusion of travel times in Section \ref{sec:CalaverasLoc-CWIandTT}
allows us to move to a global reference system.

In summary, the posterior $P(\mathbf{e}_1,...,\mathbf{e}_n |
\widetilde{\mathbf{\delta}}_{CWIN})$ and its negative logarithm $L$
describe the joint probability of multiple event locations given the
observed coda waves. The most likely set of locations is given by
the minimum of $L$.   In this paper we use the Polak-Ribiere
technique \citep{dr_Press87a}, a conjugate gradient method, to
minimize $L$. It uses the derivatives of $L$, derived in Appendix
\ref{sec-Appendix-derivatives_ofL}, to guide the
optimization procedure. Note that when optimizing
equation \ref{eq:-negative-log} the values of $ln \left[ c \right]$
and $ln \left[ P(e_i) \right]$ can be ignored because they are
constant\footnote{$ln \left[ P(e_i) \right]$ is constant because we
consider a uniform prior.}

@@

The local coordinate system works by constraining the location of
the first three earthquakes. Earthquake 1 is fixed at the origin,
earthquake 2 on the positive $\hat{x}$-axis and earthquake 3 has
$\hat{y}>0$. As the number of events increase the strength of these
constraints on later events weakens allowing small rotations of
events with respect to each other. That is, even though the
rotational freedom of the cluster is in principal removed by the
constraints imposed on the events (see equations
(\ref{eq:coord-const1}) to (\ref{eq:coord-const3})\footnote{equation
\ref{eq:coord-const4} is needed in 3D only}) we observe that in
practice the presence of noise allows the rotational non-uniqueness
to reappear. This is because errors align themselves in directions 
least constrained by data. For the CWI technique this ammounts to rotations in 2D. 
The same phenomena is observed in linear inversion where noise
creates large spurious model changes in directions of the
eigenvectors with the smallest singular values \citep{dr_Aster05a}.
 Fortunately however, combining coda waves with
measurements of travel times alleviates this problem and facilitate
the removal of a local coordinate system altogether (see Section
\ref{sec:CalaverasLoc-CWIandTT}).

@@

Figures \ref{fig:optimisationresults-2Dsynth}(a) and (b) illustrates
the maximum $\Delta_{max}$ (top) and mean $\Delta_\mu$ (middle) of
the coordinate error as a function of percentage of earthquake pairs 
that are directly linked by a separation estimate. We show
the statistics for the `best' optimization solution (black) and for
the solution space when all 25 optimizations are considered (gray).
In the latter case the best solution is determined by the set of
event locations which lead to the smallest value of $L$. The error
in the best solution is consistent when 30\% or more of the branches
are used. The errors increase when only 10\% or 20\% of the
constraints are included. Interestingly, this breakdown around 20\%
to 30\% coincides with the point where the average number of
branches required to link each event pair reaches 2 (see
Fig.\ref{fig:optimisationresults-2Dsynth} (bottom)). Since the
average number of branches can be computed in advance it can be used
as an indication of the inversion stability prior to optimization. A
higher breakdown is observed when all 25 solutions are considered
collectively. For example, the maximum coordinate error
$\Delta_{max}$ exceeds that for the best solution for linkage
$\leq$60\% confirming that the optimization is susceptible to local
minima and that a range of starting points should be considered.
Some optimizations fail to converge after 1200 iterations when the
linkage is 60\% or lower. All optimizations fail when the linkage
is 20\% or lower. Despite their failure to converge, the locations
at final iteration are close to the actual solution.

The derivatives used in the conjugate gradient method depend on
events connected by CWI measurements. Consequently, earthquakes that
are only connected via other events do not `communicate' with each
other directly. To some extent, this should be adressed during the
iterative process where location information can spread to events
which have no direct links. However, the lack of direct connection
through the gradient could prevent convergence in extreme cases, or
more likely slow the procedure down. This could explain why some
examples do not converge after 1200 iterations.
\citet{dr_VanDecar94a} show that derivative based regularization acts slowly
through iterative least-squares, because
 every cell in one iteration communicates only with its neighbours, and they demonstrate that this can be
fixed with preconditioning in some cases. Their findings suggest
that it may be possible to improve the convergence (stability and/or
speed) of the CWI optimization by preconditioning.

@@

When 70\% of the direct constraints are considered all optimization
results (gray) are consistent with the best solution obtained from 
all 25 starting locations (black). The
best solution constrains the event locations down to 30\% of the
direct links. There is one notable difference between the 3D and 2D results. In 2D
the final iteration was close to the actual solution when the
optimization failed to converge. Conversely, in 3D the optimization
appears to converge to the correct solution or fail completely,
leading to a set of locations at final iteration which do not
resemble the actual solution. This is depicted in Figure
\ref{fig:optimisationresults-3Dsynth} by the absence of the gray and
black lines below 60\% and 30\% of the constraints, respectively.
The reason for this difference may be due to the increased number of
degrees of freedom in 3D requiring a greater number of iterations to
converge. Nevertheless, the accurate convergence of the best
solution for cases with 30\% linkage or higher is encouraging for
the potential of coda wave optimization to constrain earthquake
location.

@@

In summary, the synthetic examples demonstrate the ability of coda
wave data to constrain relative event location using optimization.
The optimization error is influenced by the noise on CWI estimates
with greater $\bar{\sigma}_N$ leading to larger errors in the
solutions. When 70\% or more of the direct branches are used the
optimizer is stable with no observable difference in the solution
for 25 randomly chosen starting locations. As the direct linkage
reduces to 50\% the optimization becomes less stable and the best
solution from 25 random starting locations is required to find the
optimal solution. All optimisations fail to converge as the number 
of links decrease below 30\%.

@@

\begin{figure}
\noindent\includegraphics[width =
20pc]{diags/CalaverasMap/gmt_california/CaliforniaCalaverasMap1.eps}
\caption{Map showing location of the Calaveras
cluster (white star) and 805 seismic stations (black triangles).}
\label{fig:-eqopti-California-Calaveras}
\end{figure}

@@

Absolute locations cannot be found by CWI alone. For the sake of
comparison, we arbitrarily choose a `master' event and translate our
relative locations to align with the hypoDD location for the same
event. This arbitrary translation does not change the relative
locations. We return to this issue of relative versus absolute
location in Example 7 by introducing a combined travel time and coda
wave inversion.

@@

% This is the version using hypoDD with SVD
\begin{figure}
\noindent\includegraphics[width =
20pc]{diags/CalaverasLoc4_hypoDD_SVD.eps} \caption{Example 6 -
Number of constrainable events,$nE$ in
the CWI and hypoDD inversions as a function of the stations
considered (top). Mean (middle) and maximum (bottom) of the difference
computed between the reduced station inversion results (CWI and
hypoDD) and the complete hypoDD locations for all 308 events. } \label{fig-statremoval_summarystats}
\end{figure}
%%=========================================================================

The hypoDD algorithm also fails to locate all events as the number
of stations is reduced. In the case of hypoDD an event can be
identified as unconstrainable in one of two stages. Firstly, the
data are analyzed to ensure that there exists travel time
differences for each event and at least one other earthquake. This
is analogous to the situation for the coda wave technique. The
hypoDD program also has a secondary identification phase in which
events that can not be located sufficiently are rejected during the
inversion. This process is related to the iterative
removal of outliers described by \citet{dr_Waldhauser00a}. The top
panel of Figure \ref{fig-statremoval_summarystats} shows that the
number of events re-located by hypoDD fluctuates between 63 and 28
earthquakes for 10 to 4 stations and it demonstrates that the number
of events located by hypoDD is less than the number located by CWI
in all cases.
